Sperm should evolve to make female meiosis fair.

### 2014-07-25 writeup_and_notes/GeneticsResub/male_dependent_drive.tex
added refs and small changes to intro

\yb{Before doing so, it is worth considering whether there is any evidence that sperm could (or do), influence female meiosis. }
\gc{It is becoming
        increasingly clear that sperm bring a wide variety of RNA and proteins into the
        egg REF. 
 Some of these have known functions, for example, in most
        animal species sperm-- not eggs-- are
        responsible for the transmission of the centriole, a vital
        component of the mitotic machinery for the zygote. 
Detailed functional studies and paternal effect mutations in model
systems further highlight
 that sperm-transmitted products have a wide-range of functions in
 the completion of syngamy, zygotic development, and the resumption of
 female meiosis \cite{Miller:01, McNally:05,REF}. 
 For example, in \emph{C. elegans}, premature deployment of the aster 
 	-- a vital component of mitotic machinery provided by the sperm, 
	disrupts MII meiotic segregation in the egg, leading to a triploid zygote \citep{McNally2012}.
However,
        the function of many of the products the sperm brings into the
        egg is completely unknown and 
      these products vary widely over species \citep{Karrbook:09}. 
It seems quite plausible that sperm-based products and
hence alleles could potentially have functional effects on
various aspects of female meiosis that occur after fertilization. } 
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous? 
\gc{THERE may be too much bio. detail in previous paragraph, I can
  add more refs. once we've decided on content
centriole : http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3144413/?report=classic}
\newline

%It is certainly not the case that sperm is mechanistically incapable of influencing female meiosis.
%Both mechanistic and genetic evidence suggests that sperm can influence the outcome of female meiosis.
%An example of mechanistic evidence for sperm's influence on female meiosis comes from \emph{C. elegans}, where premature deployment of the aster -- a vital component of mitotic machinery provided by the sperm, 
%	disrupts MII meiotic segregation in the egg, leading to a triploid zygote \citep{McNally2012}.
\yb{Current evidence suggests that sperm influence on female meiotic drive is not only possible, but likely. }
Evidence from both the \emph{In} and \emph{Om}  -- suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
%There are current only two polymorphic, well-characterized female meiotic drive systems in
%animals-- and both of them, the \emph{In} and \emph{Om}  drive systems in mice-- suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages 52-54 in][ for comment]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the 	
	genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}. 
\yb{The fact that the two best characterized, polymorphic systems of putative female 
	meiotic drive systems identified to date in animals (\emph{In} and \emph{Om}) 
	show this effect suggests that if female meiotic drive is
	common the role of sperm in modifying female drive will be important as well. } \newline

@@

\yb{Because female meiotic drive is likely common and consequential, 
	and because sperm have ample opportunity to influence female drive,  
	we develop population genetic models exploring the evolution of sperm influence on female meiotic drive -- 
	that is, what is the expected influence of sperm on female drive? } 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and
maternal  genomes' are often aligned, 
as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Thus, there is little selective benefit to females in preventing sperm to influence female meioses. 
 \yb{In fact, a mutant female which completes meiosis before 
 	fertilization is giving up the protection from meiotic drivers potentially offered by sperm, 
	and is therefore more likely to produce low fitness offspring. }
%\gc{We do not claim that female meiotic arrest evolved
%  specifically to recruit helpful sperm to defend against meiotic
%  drive, the delayed completion of female meiotic drive will not in
%  fact encourage sperm to assist renegade drivers. Rather, we think
%  that features of
%  o\"{o}genesis  may evolve to facilitate an interaction with sperm 
%with female meiosis
%as yet another mechanism to enforce an equitable transfer of genetic material.}

@@

\gc{Our theory also encourages phylogenetic hypotheses concerning the relationship between the opportunity 
	for female meiotic drive and the requirement of fertilization for the completion of female meiosis. 
For example, we predict that a lower opportunity for female meiotic drive,
        e.g. an animal lineage with a history of high inbreeding
        or selfing, 
        may be accompanied by a relaxation of the requirement of fertilization for the
        completion of female meiosis (although 
   \yb{     opportunities to test this hypothesis may be limited because }
        %this effect may be weaks as such
        lineages may only persist for a short time). 
Our results also suggests that 
%Another   hypothesis that our work suggests is we might expect 
        that phylogenetic variation in the stage of female meiosis when
        fertilization occurs to influence the prevalence of
        female meiotic drive.
For example, in taxa where meiosis I is completed
by the egg pre-fertilization, including humans, centromere drive may
be more common as sperm based modifiers can not intercede than in
species where fertilization occurs during MI. 
Such hypotheses could be tested if the speed of centromere
turnover serves as a proxy for the efficacy of MI drive. }
\newline

>>>>

Before doing so, it is worth considering whether there is any evidence that sperm could (or do), influence female meiosis. 
It is becoming
        increasingly clear that sperm bring a wide variety of RNA and proteins into the
        egg \citep{Miller:05}. 
 Some of these have known functions, for example, in most
        animal species sperm-- not eggs-- are
        responsible for the transmission of the centriole, a vital
        component of the mitotic machinery for the zygote \citep{Schatten:94}. 
Detailed functional studies and paternal effect mutations in model
systems further highlight
 that sperm-transmitted products have a wide-range of functions in
 the \gc{egg activation, completion of syngamy, zygotic development, and the resumption
 and successful completion of
 female meiosis} \cite[e.g.][]{Yasuda:95,Loppin:05,Miller:01, McNally:05,Churchill:03}. 
 For example, in \emph{C. elegans}, \gc{premature deployment of the sperm aster 
	disrupts MII meiotic segregation in the egg, leading to a triploid zygote} \citep{McNally2012}.
However,
        the function of many of the products the sperm brings into the
        egg is completely unknown and 
      these products vary widely over species \citep{Karrbook:09}. 
It seems quite plausible that sperm-based products and
hence alleles could potentially have functional effects on
various aspects of female meiosis that occur after fertilization. 
%Why then is the requirement of fertilization to complete female meiosis so ubiquitous? 
\newline

%It is certainly not the case that sperm is mechanistically incapable of influencing female meiosis.
%Both mechanistic and genetic evidence suggests that sperm can influence the outcome of female meiosis.
%An example of mechanistic evidence for sperm's influence on female meiosis comes from \emph{C. elegans}, where premature deployment of the aster -- a vital component of mitotic machinery provided by the sperm, 
%	disrupts MII meiotic segregation in the egg, leading to a triploid zygote \citep{McNally2012}.
Current evidence suggests that sperm influence on female meiotic drive is not only possible, but likely. 
Evidence from both the \emph{In} and \emph{Om}  -- suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
%There are current only two polymorphic, well-characterized female meiotic drive systems in
%animals-- and both of them, the \emph{In} and \emph{Om}  drive systems in mice-- suggests that the 
	 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages 52-54 in][ for comment]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the 	
	genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}. 
\yb{The fact that the two best characterized, polymorphic systems of putative female 
	meiotic drive systems identified to date in animals (\emph{In} and \emph{Om}) 
	show this effect suggests that if female meiotic drive is
	common the role of sperm in modifying female drive will be important as well. } \newline

@@

\yb{Because female meiotic drive is likely common and consequential, 
	and because sperm have ample opportunity to influence female drive,  
	we develop population genetic models exploring the evolution of sperm influence on female meiotic drive -- 
	that is, what is the expected influence of sperm on female drive? } 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and
maternal  genomes' are often aligned, 
as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Thus, there is little selective benefit to females in preventing sperm to influence female meioses. 
\gc{Given the wide-spread requirement of fertilization for the completion
of  female meiosis it seems possible that various feature of the
interaction between sperm and egg have evolved as another mechanism to enforce an equitable transfer of genetic material.}
\gc{HOW ABOUT THE ABOVE.}

%%don't thi k this helped: 
%\yb{In fact, a mutant female which completes meiosis before 
 %	fertilization is giving up the protection from meiotic drivers potentially offered by sperm, 
%	and is therefore more likely to produce low fitness offspring. }
%%\gc{We do not claim that female meiotic arrest evolved
%  specifically to recruit helpful sperm to defend against meiotic
%  drive, the delayed completion of female meiotic drive will not in
%  fact encourage sperm to assist renegade drivers. Rather, we think
%  that features of
%  o\"{o}genesis  may evolve to facilitate an interaction with sperm 
%with female meiosis
%as yet another mechanism to enforce an equitable transfer of genetic material.}

@@

\gc{Our theory also encourages phylogenetic hypotheses concerning the relationship between the opportunity 
	for female meiotic drive and the requirement of fertilization for the completion of female meiosis. 
For example, we predict that a lower opportunity for female meiotic drive,
        e.g. an animal lineage with a history of high inbreeding
        or selfing, 
        may be accompanied by a relaxation of the requirement of fertilization for the
        completion of female meiosis (although 
   \yb{     opportunities to test this hypothesis may be limited because }
        %this effect may be weaks as such
        lineages may only persist for a short time). 
Our results also suggests that 
%Another   hypothesis that our work suggests is we might expect 
        that phylogenetic variation in the stage of female meiosis when
        fertilization occurs may influence the prevalence of
        female meiotic drive.
For example, in taxa where meiosis I is completed
by the egg pre-fertilization, including humans, centromere drive may
be more common as sperm based modifiers can not intercede than in
species where fertilization occurs during MI. 
Such hypotheses could be tested if the speed of centromere
turnover serves as a proxy for the efficacy of MI drive. }
\newline
